---
title: "ONS COVID Infection Survey - Example Solution"
output:
  html_document:
    df_print: paged
---

```{r echo = FALSE}
library(tidyverse)

## ggplot2 extension packages
library(ggthemes)
library(ggalt)
```

## ONS COVID Infection Survey Dataset
This data shows the daily COVID infection rate over a recent few weeks in each of none English Regions. The central estimate and 95% credible upper and lower bounds are given.  This data is updated frequently  on the [ONS website Coronavirus (COVID-19) Infection Survey page](https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/conditionsanddiseases/bulletins/coronaviruscovid19infectionsurveypilot/19february2021).

```{r}
covid_url <-
  "https://raw.githubusercontent.com/MarkWilcock/R-Course/main/Datasets/Covid%20Infection%20Survey.csv"
covid_df <- read_csv(covid_url)
head(covid_df)
```

Let's reshape this dataset.

```{r}
covid_tidy <-
  covid_df %>%
  gather(key = "RegionMeasure", value = "Rate",-Date) %>%
  separate(RegionMeasure, c("Region", "Measure"), sep = "-")

covid_tidy
```


*Exercise*
Create a line chart to see how the COVID infection rate changes over time for each region
```{r}
covid_tidy_central <- covid_tidy %>% filter(Measure == "Central") %>% select(-Measure)

ggplot(covid_tidy_central,
       aes(x = Date, y = Rate, col = Region)) +
  geom_line()
```

*Exercise* 
Create a line chart that shows just the London Region - the value and the upper and lower bounds

We need these dataset for the next exercise
```{r}
covid_average_by_region <-
covid_tidy_central %>% 
  group_by(Region) %>% 
  summarise(AverageRate = mean(Rate))
```


*Exercise* 
Create a column chart that shows the average rate over the period by region
```{r}
ggplot(
covid_average_by_region,
aes(x=Region, y=AverageRate)) + 
  geom_col()
```
*Exercise* - make this chart more appealing
```{r}
library(scales) # required for the labels = percent
ggplot(covid_average_by_region,
       aes(x = Region, y = AverageRate)) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  scale_x_discrete(NULL) +
  scale_y_continuous("Average Rate In the Period", labels=percent)
```
*Exercise* Create a bar chart to show expected lifespan of women born in 2010 born by country.



## Scatter Plot (geom_point())
Scatter plots compare items across two continuous axes


# Small multiples

Let's look at the infection rate in each region

```{r}
ggplot(covid_tidy_central,
       aes(x = Date, y = Rate)) +
  geom_line() +
  facet_wrap( ~ Region)
```

With small multiples we can put more details on to each chart e.g. the upper and lower bands as well as a central band

**Exercise**: Create a small multiple chart of each region where each chart shows the central, upper and lower bands



## Improve the look and effectiveness of our charts


```{r}
ggplot(covid_tidy_central,
       aes(x = Date, y = Rate, col = Region)) +
  geom_line() +
  ggtitle(
    "The percentage of people testing positive has decreased in all regions in the week ending 12 February 2021", 
    "Estimated percentage of the population testing positive for the coronavirus (COVID-19) on nose and throat swabs, daily, by region since 2 January 2021, England")

```
labs() is possibly a better alternative since we can label title, axes and *captions* in one fell swoop.
```{r}
ggplot(covid_tidy_central,
       aes(x = Date, y = Rate, col = Region)) +
  geom_line() +
  labs(
    title = "COVID Rate decreased in all regions", 
    subtitle  = "Percentage of the population testing positive", 
    caption = "Source: https://www.ons.gov.uk/peoplepopulationandcommunity/"
  )
```



```{r}
ggplot(covid_tidy_central,
       aes(x = Date, y = Rate, col = Region)) +
  geom_line() +
  ggtitle(
    "COVID Rate decreased in all regions", 
    "Percentage of the population testing positive") +
  xlab("Date (Daily Measurements") +
  ylab("Estimated Infection Rate")
```


```{r}
ggplot(covid_tidy_central,
       aes(x = Date, y = Rate, col = Region)) +
  geom_line() +
  #theme(legend.position = "bottom") +
  scale_color_discrete (guide = guide_legend(title = "English Region"))

```
*Exercise* Move the legend to the bottom of the chart

```{r}
ggplot(covid_tidy_central,
       aes(x = Date, y = Rate, col = Region)) +
  geom_line() +
  scale_x_date ("Date Axis", date_labels = "%b %y") +
  scale_y_continuous("Infection Rate", limits = c(0, 0.04), labels = scales::percent)
```

## Reference Lines, Annotations and Labels


```{r}
(covid_tidy_central_london <-
  covid_tidy_central %>% filter(Region == "London") %>%  
  select(-Region)
)
```

```{r}
ggplot(covid_tidy_central_london    ,
       aes(x = Date, y = Rate)) +
  geom_line() +
  geom_hline(yintercept = 0.02, col="red", linetype="dashed") +
  annotate("text", x = min(covid_tidy_central_london$Date), y = 0.02, label = "Lockdown Level", vjust = -0.5, hjust = 0)

```



Try out different themes.  Use the auto-complete and documentation to see the themes available.  Which do you prefer?
```{r}
ggplot(covid_tidy_central,
       aes(x = Date, y = Rate, col = Region)) +
  geom_line() +
  ggtitle(
    "COVID Rate decreased in all regions", 
    "Percentage of the population testing positive") +
  xlab("Date (Daily Measurements") +
  ylab("Estimated Infection Rate") +
  theme_light()

```


```{r}
ggplot(covid_tidy_central,
       aes(x = Date, y = Rate, col = Region)) +
  geom_line() +
  ggtitle(
    "COVID Rate decreased in all regions", 
    "Percentage of the population testing positive") +
  xlab("Date (Daily Measurements") +
  ylab("Estimated Infection Rate") +
  theme_light() +
  theme(plot.background = element_rect(fill="yellow"))
```
*Exercise* Build a chart with a minimal themes apart from a green panel background


```{r}
ggplot(covid_tidy_central,
       aes(x = Date, y = Rate, col = Region)) +
  geom_line() +
  ggtitle(
    "COVID Rate decreased in all regions", 
    "Percentage of the population testing positive") +
  xlab("Date (Daily Measurements") +
  ylab("Estimated Infection Rate") +
  theme_economist()
  
```



**Exercise**: Create a small multiple chart of each region where each chart shows the central, upper and lower bands

```{r}
#  Exercise answer
ggplot(covid_tidy,
       aes(x = Date, y = Rate, col = Measure)) +
  geom_line() +
  facet_wrap( ~ Region)
```
